---
title: "Predoc Data Task: Data Manipulation"
date: 2023-09-28
categories: ["rlang"]
---

This is my solution to the sample data manipulation task published by [PreDoc.org](https://predoc.org). 

You can find the full task [here](https://predoc.org/how-to-prepare). This is exercise 2 from data task 1.

The task is to load some health insurance subscription data, manipulate the data to develop a summary of certain variables by county (rather than by insurance plan), and then output that data in a specified format.

# Load libraries.

```{r}
#| output: false
library("dplyr")
library("readr")
library("tidyr")
```

# Read Data

```{r}
#| output: false
medicare <- read_csv("./task/scp-1205.csv", col_names = FALSE)

# `contract` is missing from the list of variables given in the instructions
# it is included with the name `contract` in the sample row
names(medicare) <- c(
  "countyname",
  "state",
  "contract",
  "healthplanname",
  "typeofplan",
  "countyssa",
  "eligibles",
  "enrollees",
  "penetration",
  "ABrate"
)
```

# Fill Missing Values

```{r}
medicare_filled_missing <- medicare |>
  mutate(
    across(
      c(eligibles, enrollees, penetration),
      ~ replace_na(.x, 0)
    )
  )
```

# Complete Objective

```{r}
medicare_by_county <- medicare_filled_missing |>
  # Filter territories and District of Columbia
  filter(!state %in% c("PR", "GU", "DC", "VI", "AS")) |>
  # Filter unspecified counties
  filter(!(is.na(state) | state == 99)) |>
  group_by(countyname) |>
  summarize(
    # first() is used where all the counties will have the same value
    state = first(state),
    numberofplans1 = sum(enrollees > 10),
    numberofplans2 = sum(penetration > .5),
    countyssa = first(countyssa),
    eligibles = first(eligibles),
    totalenrollees = sum(enrollees),
    totalpenetration = 100 * totalenrollees / eligibles
  ) |>
  arrange(state, countyname)
```

# Write to CSV

```{r}
# Commented out for the web writeup.
# write_csv(medicare_by_county, "medicare_output.csv")

# Instead print some of the data.
head(medicare_by_county, 100)
```
